Final paper: Towards Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-05-02 paper/paper.tex
paper: write about Coq

All code may be found at the project repository~\cite{zelda_mosaic_proof}. Most
of the work in this section corresponds to the file \texttt{Lib Matrix.v}.

@@

>>>>

All code may be found at the project repository~\cite{zelda_mosaic_proof}. Most
of the work in this section corresponds to the file \texttt{LibMatrix.v}.

@@

There are three major advanced Coq features that assisted our development of the
matrix library. These are
\begin{inlist}
\item the use of Ltac to simplify and automate proofs;
\item the use of hint databases to enhance the \texttt{auto} proof-search
    tactic; and
\item the use of specialized induction principles to enhance induction with
    necessary information.
\end{inlist}
We cover each in turn.

\subsubsection{Automating Coq via Ltac}\label{S:coq_ltac}

Ltac is one of Coq's languages for defining custom tactics. It can be used for
simple abbreviations (\textit{e.g.}, \texttt{gen} for \texttt{generalize dependent}) and
complex proof machinery (\textit{e.g.}, find two equalities over the same
variable that contradict each other and exploit the contradiction). We use Ltac
to enhance several standard Coq details by use of the file \texttt{LibMatrix.v}
from \textit{Programming Language Foundations}~\cite[Chs.\@ \textit{A Collection
of Handy General-Purpose Tactics} and \textit{A Gentle
Introduction}]{Pierce:SF2}. This includes tactics like \texttt{introv} for
introducing variables automatically and \texttt{inverts} for performing the
usually necessary substitutions from \texttt{inversion} on a hypothesis.

We also use Ltac to abbreviate a custom induction principle (see
\S~\ref{S:coq_ind}) and to automate proofs of the computational well-formed
property in concrete cases. The custom tactic \texttt{wf\_easy} can prove that
all of the well-formed examples given in \S~\ref{S:matrix_defn} are indeed
well-formed. It is particularly useful to solve base cases in proofs. The
\texttt{wf\_easy} tactic unfolds a few relevant definitions before repeatedly
applying a strategy to prove the recursive conjuncts generated by the
computation definition.

\subsubsection{Guiding proof search with hint databases}\label{S:coq_hint}

The \texttt{auto} tactic searches for proof chains in the hypotheses and a core
database of facts called ``hints.'' We can provide new databases full of
domain-specific hints; \texttt{auto with hints} will instruct the \texttt{auto}
tactic to search \texttt{hints} in addition to the hypotheses and the core
database.

We create a hint database \texttt{matrix} and add the relevant constructors,
definitions, and theorems to it as they are defined and proved. This allows
subsequent proofs to benefit from \texttt{auto with matrix} to solve goals that
use properties of the matrices. In particular, this helps us focus on the
conceptual challenges and connections of the proof rather than on the details of
searching for the right chain of hypotheses and theorems. Because this hint
database is exported, consumers of the library can similarly benefit.

\subsubsection{Enhancing induction through proof computation of induction
principles}\label{S:coq_ind}

As we learned from the chapter \textit{Induction Principles} of \textit{Logical
Foundations}~\cite{Pierce:SF1}, Coq permits us to develop custom induction
principles that can change the way induction works. For example, it may be
useful to do induction on the length of a list or skipping by twos in the
naturals, rather than the standard induction on the list itself or by only
successor in the naturals. As long as we can provide a function with the correct
form that computes a proof of a general proposition \(P\) over values of the
type with certain assumed hypotheses (the base and inductive cases), we can use
\texttt{induction x using IP} to perform induction on \(x\) with
induction principle \texttt{IP}.

In our case, the induction principle for the type \(\func{matrix\_content}\) was
not powerful enough---the apparently recursive definition of \(\func{Matrix}\),
which contains a list of \(\func{matrix\_content}\)s, does not come with an
induction hypothesis that all the elements of the list obey the property under
induction. However, we can define an induction principle (called
\texttt{matrix\_content\_ind\_strong} in our development) which, given a proof
for the scalar base case and a proof for the recursive matrix case \emph{under
the assumption that all the elements of the list also satisfy the property},
correctly computes \(\forall A, (m: \func{matrix\_content}\, A) : P\, m\). It does
so using list induction under the hood.

We provide the \texttt{mc\_ind} tactic as shorthand for induction using the
above principle. We use it in most proofs by induction on the contents of a
matrix. Because the induction hypothesis is strengthened, however, we often have
to make it more general (\textit{e.g.}, by generalizing the shape of the matrix)
in order to successfully complete the proof.
